Evaluating Cloud-Optimized HDF5 for NASA’s ICESat-2 Mission

### 2024-10-09 paper.qmd
fixing typos

The most recent version of the Hierarchical Data Format is HDF5, a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between. As flexible and powerful as HDF5 can be, it comes with big tradeoffs when it’s accessed from remote storage systems.

@@

The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”. Page aggregation does 2 things: fisrt, it collects file-level metadata from datasets and stores it on dedicated metadata blocks at the front of the file; second, it forces the library to write both data and metadata using these fixed-size pages. Aggregation allows client libraries to read  file metadata with only a few requests using the page size as a fixed request size, overriding the 1 request per chunk behavior.

@@

As we can see in [@fig-2], when we cloud optimize a file using paged-aggregation there are some considerations and behavior that we had to take into account. The first thing to observe is that 
page aggregation will --as we mentioned-- consolidate the file-evel metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file. It stores important metadata such as the version of the file format, pointers to the root group, and addresses for locating different file components]
The next thing to notice is that page size us uses across the board for metadata and data as of October 2024 and version 1.14 of the HDF5 library, page size cannot dynamically adjust to the total metadata size.

@@

This one page size for all approach simplifies how the HDF5 API reads the file (if instructucted) but it also brings unused page space and chunk over-reads. In the case of the ICESat-2 dataset (ATL03) the data itself has been partitioned and each granule
represents a segment in the satellite orbit and within the file the most relevant dataset is chunked using 10,000 items per chunk, with data being float-32 and using a fast compression value, the resulting chunk size is on average under 40KB, which is really small 
for an HTTP request, especially when we have to read them sequentially. Because of these considerations, we opted for testing different page sizes, and increase chunk size. The following table describes the different configurations used in our tests.

@@

shows that using paged aggregation alone is not a complete solution. This behavior us caused by over-reads of data now distributed in pages and the internals of HDF5 not knowing how to optimize 
the requests. This means that if we cloud optimize alone and use the same code, in some cases we'll make access to these files even slower. A very important thing to notice here is that rechunking the file, in this case using 10X bigger chunks results in a predictable 10X improvement in access times without any cloud optimization involved.
Having less chunks generates less metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have anough memory and bandwith even
bigger (Pangeo recommends up to 100MB chunks)[Add citation.]

@@

1. File page strategy: H5F_FSPACE_STRATEGY_PAGE
2. File page size: 8000000
  If repacking an existing file, h5repack contains the code to alter tese variables inside the file 
  ```bash
   h5repack -S PAGE -G 8000000 input.h5 output.h5
  ```
3. Avoid using unlimited dimensions when creating variables because the HDF5 API cannot support it inside bffered pages and representation of these variables is not supported by Kerchunk.

>>>>

The most recent version of the Hierarchical Data Format is HDF5, a common archival format for n-dimensional scientific data; it has been utilized to store valuable information from astrophysics to earth sciences and everything in between. As flexible and powerful as HDF5 can be, it comes with big trade-offs when it’s accessed from remote storage systems.

@@

The test files were originally cloud optimized by  “repacking” them, using a relatively new feature in the HDF5 C API called “paged aggregation”. Page aggregation does 2 things: first, it collects file-level metadata from datasets and stores it on dedicated metadata blocks at the front of the file; second, it forces the library to write both data and metadata using these fixed-size pages. Aggregation allows client libraries to read  file metadata with only a few requests using the page size as a fixed request size, overriding the 1 request per chunk behavior.

@@

As we can see in [@fig-2], when we cloud optimize a file using paged-aggregation there are some considerations and behavior that we had to take into account. The first thing to observe is that 
page aggregation will --as we mentioned-- consolidate the file-level metadata at the front of the file and will add information in the so-called superblock^[The HDF5 superblock is a crucial component of the HDF5 file format, acting as the starting point for accessing all data within the file. It stores important metadata such as the version of the file format, pointers to the root group, and addresses for locating different file components]
The next thing to notice is that page size us uses across the board for metadata and data as of October 2024 and version 1.14 of the HDF5 library, page size cannot dynamically adjust to the total metadata size.

@@

This one page size for all approach simplifies how the HDF5 API reads the file (if configured) but it also brings unused page space and chunk over-reads. In the case of the ICESat-2 dataset (ATL03) the data itself has been partitioned and each granule
represents a segment in the satellite orbit and within the file the most relevant dataset is chunked using 10,000 items per chunk, with data being float-32 and using a fast compression value, the resulting chunk size is on average under 40KB, which is really small 
for an HTTP request, especially when we have to read them sequentially. Because of these considerations, we opted for testing different page sizes, and increase chunk size. The following table describes the different configurations used in our tests.

@@

shows that using paged aggregation alone is not a complete solution. This behavior us caused by over-reads of data now distributed in pages and the internals of HDF5 not knowing how to optimize 
the requests. This means that if we cloud optimize alone and use the same code, in some cases we'll make access to these files even slower. A very important thing to notice here is that rechunking the file, in this case using 10X bigger chunks results in a predictable 10X improvement in access times without any cloud optimization involved.
Having less chunks generates less metadata and bigger requests, in general is it recommended that chunk sizes should range between 1MB and 10MB[Add citation, S3 and HDF5] and if we have enough memory and bandwidth even
bigger (Pangeo recommends up to 100MB chunks)[Add citation.]

@@

1. File page strategy: H5F_FSPACE_STRATEGY_PAGE
2. File page size: 8000000
  If repacking an existing file, h5repack contains the code to alter these variables inside the file 
  ```bash
   h5repack -S PAGE -G 8000000 input.h5 output.h5
  ```
3. Avoid using unlimited dimensions when creating variables because the HDF5 API cannot support it inside buffered pages and representation of these variables is not supported by Kerchunk.
